---
title: "lecture 17 simulation"
date: "2025-10-30"
output: github_document
---

```{r include = FALSE}
library(tidyverse)
set.seed(1)
```

### Simulation: Mean and SD for one *n*

The function from lecture 16, which simulates data from a normal distribution and returns estimates of the mean and sd for that data:

```{r}
sim_mean_sd = function(n, mu = 2, sigma = 3) {
  
  sim_data = tibble(
    x = rnorm(n, mean = mu, sd = sigma),
  )
  
  sim_data %>% 
    summarize(
      mu_hat = mean(x),
      sigma_hat = sd(x)
    )
}
```

IRL, repeated sampling is difficult and annoying. On a computer, it's pretty easy, using simulations. E.g., running the above function 100 times, to see the effective randomness of *x* and on the mean and sd estimates:

```{r}
output = vector("list", 100)

for (i in 1:100) {
  output[[i]] = sim_mean_sd(30)
}

sim_results = bind_rows(output)

sim_results
```

Replicating the above in an alternative form:

- Iterate 100 times with a fixed sample size of 30
- map `sim_mean_sd` over `sample_size` column to replicate the previous simulation

```{r}
sim_results_df = 
  expand_grid(
    sample_size = 30,
    iter = 1:100) %>% 
  mutate(
    estimate_df = map(sample_size, sim_mean_sd)) %>% 
  unnest(estimate_df)

sim_results_df
```

The end result is a dataframe that can then be manipulated:

```{r}
sim_results_df %>% 
  ggplot(aes(x = mu_hat)) + 
  geom_density() +
  theme_minimal()
```

```{r}
sim_results_df %>% 
  pivot_longer(
    mu_hat:sigma_hat,
    names_to = "parameter", 
    values_to = "estimate") %>% 
  group_by(parameter) %>% 
  summarize(
    emp_mean = mean(estimate),
    emp_sd = sd(estimate)) %>% 
  knitr::kable(digits = 3)
```

This provides us information on the distribution in a theoretical scenario, which can be compared against empirical models, and also provides a way to test statistical procedures with repeated sampling in a way that wouldn't be possible with single data sets.

In cases where function input doesn't change, using an "anonymous" function with the syntax `(i)` which defines a function with the input `i`, lets you rerun the function multiple times while keeping the input the same:

```{r}
sim_results_df =   
  map(1:100, \(i) sim_mean_sd(30, 2, 3)) %>% 
  bind_rows()

sim_results_df
```

### Simulation: Mean for several *n*s

An example of iterating several times with a fixed *set* of sample sizes:

```{r}
sim_results_df = 
  expand_grid(
    sample_size = c(30, 60, 120, 240),
    iter = 1:1000) %>% 
  mutate(
    estimate_df = map(sample_size, sim_mean_sd)) %>% 
  unnest(estimate_df)

sim_results_df
```

We can then manipulate the above data to see what it looks like:

```{r}
sim_results_df %>% 
  mutate(
    sample_size = str_c("n = ", sample_size),
    sample_size = fct_inorder(sample_size)) %>% 
  ggplot(aes(x = sample_size, y = mu_hat, fill = sample_size)) + 
  geom_violin() +
  theme_minimal()
```

```{r message = FALSE}
sim_results_df %>% 
  pivot_longer(
    mu_hat:sigma_hat,
    names_to = "parameter", 
    values_to = "estimate") %>% 
  group_by(parameter, sample_size) %>% 
  summarize(
    emp_mean = mean(estimate),
    emp_var = var(estimate)) %>% 
  knitr::kable(digits = 3)
```

### Simulation: SLR for one *n*

A function estimating intercept and beta-1 for a simulated simple linear regression:

```{r}
sim_regression = function(n, beta0 = 2, beta1 = 3) {
  
  sim_data = 
    tibble(
      x = rnorm(n, mean = 1, sd = 1),
      y = beta0 + beta1 * x + rnorm(n, 0, 1)
    )
  
  ls_fit = lm(y ~ x, data = sim_data)
  
  tibble(
    beta0_hat = coef(ls_fit)[1],
    beta1_hat = coef(ls_fit)[2]
  )
}
```

Which we can then run 500 times to show the effect of randomness on intercepts and beta-1s:

```{r}
sim_results_df = 
  expand_grid(
    sample_size = 30,
    iter = 1:500) %>% 
  mutate(
    estimate_df = map(sample_size, sim_regression)) %>% 
  unnest(estimate_df)

sim_results_df
```

We can then plot these estimates against each other on a plot, and see the overall trend in the way they are effected by randomness:

```{r}
sim_results_df %>% 
  ggplot(aes(x = beta0_hat, y = beta1_hat)) + 
  geom_point() +
  theme_minimal()
```

### Varying two simulation parameters

An example of iterating several times, where two different parameters have multiple options:

```{r}
sim_results_df = 
  expand_grid(
    sample_size = c(30, 60, 120, 240),
    true_sd = c(6, 3),
    iter = 1:1000) %>% 
  mutate(
    estimate_df = 
      map2(sample_size, true_sd, \(n, sd) sim_mean_sd(n = n, sigma = sd))) %>% 
  unnest(estimate_df)

sim_results_df
```

Which we can then plot:

```{r}
sim_results_df %>% 
  mutate(
    true_sd = str_c("True SD: ", true_sd),
    true_sd = fct_inorder(true_sd),
    sample_size = str_c("n = ", sample_size),
    sample_size = fct_inorder(sample_size)) %>% 
  ggplot(aes(x = sample_size, y = mu_hat, fill = sample_size)) + 
  geom_violin() + 
  facet_grid(. ~ true_sd) +
  theme_minimal()
```

